Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-11-22 SEIR.tex
continue edits

\todo[inline]{add structured pop refs}
To relax the assumption of a homogeneous population, the population can be stratified.
If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors (\eg genetics reducing the susceptibility of some individuals~\autocite[208]{andersonInfectious}).
I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
Stratification requires two major changes model changes: extension of the state vector and describing the between-strata interactions.

@@

The interactions between strata also need describing.
This is done by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}; its interpretation is described below.
The force of infection is now a vector, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
Its interpretation can be arrived at by considering a single infected individual in strata $a'$, analogous to the previous explanation in the single stratum case.
Assume any such individual comes into contact with individuals in strata $a$ at a rate of $\kappa_{aa'}$ in units of contacts per day.
% Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
% For such a randomly selected individual, the probability that they are susceptible is $s_a(t)$.
% Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Following the same argument as in \cref{SEIR:sec:SIR} finds that, in expectation, the infected individual makes contact with susceptible individuals in stratum $a$ at a rate of $\kappa_{aa'} s_a(t)$.
Assume that there is a constant probability of transmission upon contact between an infected in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
Therefore, the expected rate of infections in strata $a$ caused by this infected individual is $\kappa_{aa'} s_a(t) q_{aa'}$.
Multiplying by the number of infected individuals in strata $a'$, $N_{a'} i_{a'}(t)$, gives $\kappa_{aa'} q_{aa'} s_a(t) N_{a'} i_{a'}(t)$.
The constant terms here are absorbed into the WAIFW matrix, giving $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$~\autocite{section 9.2}{diekmannMathematical}.
The total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$.
Equivalently, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
The rate that individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.

In practical situations, estimating all of the $\beta_{aa'}$ parameters is infeasible, so some assumptions must be made to reduce the number of parameters~\autocite[176]{andersonInfectious}.
It is infeasible because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
Common assumptions include assuming symmetry, \ie $\beta_{aa'} = \beta_{a'a}$, or that the direction of transmission is irrelevant, \ie $\beta_{aa'} = \beta_{a} \beta_{a'}$.
% Alternatively, as in the scalar case, it may be easier to iden $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
The choice of assumptions depends on the context, including the disease being studied and the data available.

@@

The modelled proportion of individuals that are PCR positive in the can then be linked to the data using a Beta-Binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \sum_{k=1}^l \bar{p}_{kat}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
In this parameterisation (see \cref{E-distributions}) at $\rho=0$ the beta-binomial coincides with the binomial distribution and as $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$.

@@

\todo[inline]{sort region names in plots}

@@

\begin{figure}
    \includegraphics{SEIR/CIS/p_peak}
    \caption[Posterior estimates of peak incidence timing]{%
        Posterior estimates (95\% CrI) of incidence by region and age.
        See main text for discussion.
    }
    \label{SEIR:fig:incidence}
    \todo[inline]{write peak incidence caption}
\end{figure}

>>>>

To relax the assumption of a homogeneous population, the population can be stratified.
If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors (\eg genetics reducing the susceptibility of some individuals~\autocite[208]{andersonInfectious}).
I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
Stratification requires two major changes model changes: extension of the state vector and describing the between-strata interactions.

@@

The interactions between strata also need describing.
This is done by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}; its interpretation is described below.
The force of infection is now a vector, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
Its interpretation can be arrived at by considering a single infected individual in strata $a'$, analogous to the previous explanation in the single stratum case.
Assume any such individual comes into contact with individuals in strata $a$ at a rate of $\kappa_{aa'}$ in units of contacts per day.
% Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
% For such a randomly selected individual, the probability that they are susceptible is $s_a(t)$.
% Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Following the same argument as in \cref{SEIR:sec:SIR} finds that, in expectation, the infected individual makes contact with susceptible individuals in stratum $a$ at a rate of $\kappa_{aa'} s_a(t)$.
Assume that there is a constant probability of transmission upon contact between an infected in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
Therefore, the expected rate of infections in strata $a$ caused by this infected individual is $\kappa_{aa'} s_a(t) q_{aa'}$.
Multiplying by the number of infected individuals in strata $a'$, $N_{a'} i_{a'}(t)$, gives $\kappa_{aa'} q_{aa'} s_a(t) N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
The constant terms here are absorbed into the WAIFW matrix, giving $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
The total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$.
Equivalently, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
The rate that individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.

In practical situations, estimating all the $\beta_{aa'}$ parameters is infeasible, so some assumptions must be made to reduce the number of parameters~\autocite[176]{andersonInfectious}.
It is infeasible because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
Common assumptions include assuming symmetry, \ie $\beta_{aa'} = \beta_{a'a}$, or that the direction of transmission is irrelevant, \ie $\beta_{aa'} = \beta_{a} \beta_{a'}$.
% Alternatively, as in the scalar case, it may be easier to iden $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
The choice of assumptions depends on the context, including the disease being studied and the data available.

@@

The modelled proportion of individuals that are PCR positive in the can then be linked to the data using a Beta-Binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \sum_{k=1}^l \bar{p}_{kat}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
In this parameterisation (detailed in \cref{E-distributions}) at $\rho=0$ the beta-binomial coincides with the binomial distribution and as $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.

@@

\todo[inline]{format region names in plots}

@@

\begin{figure}
    \includegraphics{SEIR/CIS/p_peak}
    \caption[Posterior estimates of peak incidence timing]{%
        Posterior probability of the peak incidence for each region and age combination occurring on each day.
    }
    \label{SEIR:fig:peak-incidence}
\end{figure}
